Best Practices for Multiple Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-09-04 paper/manuscript.tex

\paragraph{Perturbation maps}
\tocsectioncomment{Perturbation maps paragraph: @Mary Pitman/ @Jenke Scheen/ OpenFE team to update discussion of perturbation maps?}
Based on the input ligand series, a perturbation map or network can be planned. Recent heuristics have shown the more connected the perturbation network the better. However, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}. Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or to deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted. Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig6_types_of_networks/Figure.pdf}
    \caption{\textbf{Examples of perturbation networks} (\textbf{A}) Star shaped network with the crystal structure in the center. (\textbf{B}) Network with cycle closures (see more on this in Sec.~\ref{sec:are-they-good}). Arrows indicate the direction of the perturbation. Fully converged binding free energy calculations yield binding free energy changes which sum to zero around any closed cycle. However, in practice errors may not sum to zero around closed cycles, providing a way to look for potential sampling problems. Here in (B), green cycles indicate cycles with hypothetically good cycle closure, red those with poor cycle closure. The red arrow indicates a poorly converged simulation that would give rise to bad cycle closures. The diamond indicates the use of a crystallographic binding mode.}
    \label{fig:fig_types_of_networks}
\end{figure}

In some cases, during the lead optimization stage, or for very large datasets that would benefit from rougher initial free energy ranking, or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used. However, adding redundancy into the network means that a better error analysis can be carried out by looking at cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.

Methods from experimental design have been applied to the construction of the perturbation maps. Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance. Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations. Both approaches lead to substantial reduction in the statistical error of the estimated free energies.

@@

\subsection{Best practices for reporting data }
\label{sec:plot_data}
\tocsubsectioncomment{@OpenFE team/ @Jenke Scheen/ Volunteers to expand on utility/ limitations of common metrics. Also see @Michael Gilson's GitHub issue.}
Following best practices for data generation and their analysis does not mean that data is reported in the optimal way. As a practitioner of alchemical free energy simulations you also should use best practices for reporting and plotting your results. We encourage the following standard set of analyses and ways to represent data. 
\paragraph{Statistics to include in reporting data}
As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in free energy predictions. First, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph). We recommend performing triplicates of your predictions at minimum, with starting points that are expected to be uncorrelated, to ensure some measure of reliability in your data. This replication may seem excessive, but uncertainty estimates often underestimate the true statistical uncertainty. Where performing multiple replicas of the simulation is not possible, an error estimate from e.g. MBAR can be used, though bearing in mind this is likely an underestimated error.

@@

An example of a best practice scatter comparison between computed and experimental values is shown in Fig.~\ref{fig:scatterplot_analysis}, highlighting outliers, error bars and confidence intervals. The data for this plot is artificially generated for illustration purposes.

@@

\subsection{Unsettled practices}

@@

\tocsubsubsectioncomment{Include discussion of active learning/ fast ABFE for virtual screening?}

>>>>

In more detail, simulation planning for a standard FEC can be conceptualized as a three-component framework. Starting, the process involves identifying all mappings of the involved molecules (e.g. atom mappings), which define the structural relationships and transformations between molecule A and molecule B. Subsequently, these changes, along with their shared structural elements, are evaluated by estimating the difficulty of the transformations usually with predefined scoring functions. The final step entails constructing the perturbation graph using the scores from the prior step, which serves as a fundamental tool setting up a simulation campaign.

\paragraph{Atom mappings}
For minor molecular transformations, like in most large-scale approaches, single-coordinate set approaches are typically utilized due to their relatively low computational cost. These methods necessitate atom mappings to ensure their effective application. In large-scale implementations, the efficiency and robustness of automated mapping techniques are critical factors in these endavors.

%SIS
Subgraph isomorphism solver (SIS)-based maximum common substructure (MCSS) search methods currently represent the most widely used approach for addressing atom mapping challenges. Several cheminformatics tools implementing these solutions are readily available, including pmx~\cite{gapsys2015pmx}, LOMAP~\cite{liu2013lead}, FESetup~\cite{loeffler2015fesetup}, ProtoCaller~\cite{suruzhon2020protocaller}, and SMArt~\cite{petrov2021perturbation}.

These methods exhibit high accuracy in processing topological 2D information. However, hybrid approaches such as pmx~\cite{gapsys2015pmx} and LOMAP~\cite{liu2013lead}, which incorporate 3D structural data, enhance robustness in stereochemical considerations and molecular alignment, requiring well-prepared input coordinates. Regarding computational efficiency, SIS solvers must contend with the inherent complexity of solving an NP-complete problem~\cite{raymond2002maximum, Duesbury2018Comparison}. The use of an initial MCSS seed can mitigate this computational burden~\cite{raymond2002maximum}.

% Geometry
An alternative to SIS-based MCSS methods is geometric bipartite matching for atom mapping, as implemented in tools such as pmx~\cite{gapsys2015pmx} and Kartograf~\cite{ries2024kartograf}. These algorithms enable rapid mappings due to their reduced computational complexity ($O(N) = N^3$), provided that well-aligned input geometries are available~\cite{ries2024kartograf}, through e.g. restraint-docking or prior-alignment. To accommodate topological variations, Kartograf applies a subsequent filtering step following the 3D mapping process.

% Short discussion of robustness and speed
Both mapping approaches present distinct advantages and limitations, each associated with specific prerequisites. Optimizing the robustness, and ignoring computational effort (e.g. mitigated by parallelization), it may be beneficial to employ both methods and implement a scoring mechanism to evaluate their performance. Most mappings are expected to be highly similar, this strategy could enhance robustness in the mapping process.

In terms of computational efficiency, the choice between methods largely depends on input conditions. If molecular structures are pre-aligned, the geometric approach offers a substantial speed advantage. The magnitude of the speed advantage is dependent on the number of mappings required and the number of atoms to be considered. Here we want to exemplify the effect by using the example of \textit{Ries et al.} calculating all-to-all mappings for given sets of molecules. The first example was a small set of 14 molecules and an average $38.6$ atoms per molecule; both approaches needed only $3~s$ to calculate the all-to-all mappings. A more challenging dataset with 16 molecules and an average of $74$ atoms per molecule increased the calculation duration to $26~s$ for the SIS based approach with LOMAP and $8~s$ for the geometric approach with Kartograf. Finally, using the double amount of molecules leads to a significant difference of $58~s$ for the geometric approach and $11~min$ and $16~s$ for the SIS based approach, indicating the algorithmic complexity difference. The complexity, however, does not only scale with the number of molecules. The last example of \textit{Ries et al.}  tries to put the number of atoms to an \textit{extreme} using a protein, with a very large number of atoms ($9349$ atoms) and mutates one residue, leading to two different molecules. This challenge is usually not addressed as a whole with SIS-algorithms, but chunked down into smaller sub-problems, like mappings per residue. However, the impact of the number of atoms on the computational efficiency can be well observed here. The problem was solved with the geometric approach in $2~s$. The SIS-based approach was stopped after $4~h$. \cite{ries2024kartograf} From this example, it becomes clear that for larger datasets the mapping problem becomes comparably large and starts to grow to a significant factor even in comparison to the simulation time.

Conversely, when ligands lack predefined 3D coordinates, the SIS-based methods provide an easy solution, allowing coordinate generation at a later stage. This nuanced selection process ensures optimal performance based on the specific characteristics of the molecular system under investigation.

In combination with a suitable mapping scoring function the next steps is to build a pertubation map.

\paragraph{Perturbation maps}
For RBFE campaigns with more than two ligands in the input series a \textit{perturbation map} or \textit{-network} must be generated: this is simply a set of data that describe which transformations will be executed between the ligands (referred to as \textit{edges}). ~\cite{liu2013lead} Heuristics have shown the more connected the perturbation network the higher the free energy prediction accuracy versus experimental measures. Running a fully-connected network (i.e. simulate transformations between \textit{all} ligands in the campaign) is typically considered too computationally expensive and may not be suited optimally with current estimators such as MLE (\ref{subsec:uncertainty}).~\cite{scheen22data} Several styles of network designs have been proposed in recent years, although no single method seems to consistently outperform all others.~\cite{ries2024Konnektor} There exist several methods for optimization of network structure while minimizing the number of perturbations that need to be computed, reducing the resulting computational cost at a minimal decrease in accuracy ~\cite{yang2020optimal,xu2019optimal,Pitman2023HiMap}. Sometimes - depending on the RBFE implementation - the introduction of intermediate ligands that are not part of the original congeneric series can be helpful for avoiding challenging transformations such as ring breaking or transformations that would otherwise result in large numbers of atoms being inserted or deleted.~\cite{schoenmaker2025imerge} Some commercial tools have well-calibrated underlying heuristics to plan these network designs but may fail with complicated input datasets, thus requiring manual user validation, in particular when dealing with chiral compounds.

\begin{figure}
    \includegraphics[width=0.95\columnwidth]{figures/fig6_types_of_networks/Figure.pdf}
    \caption{\textbf{Examples of perturbation networks} (\textbf{A}) Star shaped network with the crystal structure in the center. (\textbf{B}) Network with cycle closures (see more on this in Sec.~\ref{sec:are-they-good}). Arrows indicate the direction of the perturbation. Fully converged binding free energy calculations yield binding free energy changes which sum to zero around any closed cycle, and non-closing cycles that do not equal zero to within combined error can indicate potential sampling problems or protocol issues. Here in (B), green cycles indicate cycles with hypothetically good cycle closure, red those with poor cycle closure. The red arrow indicates a poorly converged simulation that would give rise to bad cycle closures. The diamond indicates the use of a crystallographic binding mode.}
    \label{fig:fig_types_of_networks}
\end{figure}

In some cases, when dealing with for example very large datasets that would benefit from rougher initial free energy ranking or in cases where perturbations would be rather large, a star shaped network as seen in Fig.~\ref{fig:fig_types_of_networks} \textbf{A} is used. However, adding redundancy into the network means that a better error analysis can be carried out by analyzing cycle closure errors as discussed in sec.~\ref{sec:are-they-good}, with an example given in Fig.~\ref{fig:fig_types_of_networks}\textbf{B}.

Methods from experimental design have been applied to the construction of perturbation maps. Yang et al.~\cite{yang2020optimal} showed how to optimize the perturbation map by selecting a fixed number of calculations from the pairwise perturbations so that the resulting set of calculations minimize the total variance. Xu~\cite{xu2019optimal} showed how to optimize the perturbation map by allocating different amounts of simulation time to different pairwise perturbations so as to minimize the total variance, given the total simulation time of all the perturbation calculations. Both approaches lead to substantial reduction in the statistical error of the estimated free energies.

@@

\subsection{Best practices for reporting data }
\label{sec:plot_data}
\tocsubsectioncomment{@OpenFE team/ @Jenke Scheen/ Volunteers to expand on utility/ limitations of common metrics. Also see @Michael Gilson's GitHub issue.}
Following best practices for data generation and their analysis does not mean that data is reported in the optimal way. As a practitioner of alchemical free energy simulations you also should use best practices for reporting and plotting your results. We encourage the following standard set of analyses and ways to represent data. Some distinctions can be made between statistics that report how accurate and reproducible your results are to a ground-truth experiment, and how useful a method might be in an applied setting. For the case of relative free energy calculations, results might be reported either per-edge, as DDGs, or as the MLE-augmented DG results. Relative results (and outliers in this regime) can be more indicative of edge-based errors, whereas outliers in DG could be more indicative of a node-based error, such as a forcefield error for particular ligands. For relative free energy results, it is best practice to plot both.

\paragraph{Statistics to include in reporting data}
As with any modelling technique, misuse of statistical analysis can skew the perception of how well models perform in free energy predictions. First, error estimates should always be included on your predictions in whatever form you present your data (scatterplots, barplots, etc; see next paragraph). We recommend performing triplicates of your predictions at minimum, with starting points that are expected to be uncorrelated, to ensure some measure of reliability in your data. This replication may seem excessive, but uncertainty estimates often underestimate the true statistical uncertainty. Where performing multiple replicas of the simulation is not possible, an error estimate from e.g. MBAR can be used, though bearing in mind this is likely an underestimated error.

@@

Best-practices for reporting predicted affinities retrospectively have been outlined, however using free energy methods prospectively where possibly none, one or only a few experimental datapoints for the set have associated energies might be available can be more difficult if any supporting statistics will not be possible or meaningful for a small set. Lipohilic ligand effciency (LLE or LipE) plots are typically used in drug discovery for understanding the relationship between lipophilic-driven binding affinity for series of compounds. Predicted or measured LogD or LogP on the x-axis and measured potency on the y-axis, typically with diagonal y=x+C lines used to indicate where compounds are achieving better/worse binding due to specific or polar interactions. Using \textit{in silico} iLLE plots, with predicted affinities and predicted LogP or LogD can be helpful to visually present free energy results with limited experimental data.

An example of a best practice scatter comparison between computed and experimental values is shown in Fig.~\ref{fig:scatterplot_analysis}A-C for relative, absolute and LogP reporting, respectively. These highlight outliers, error bars and confidence intervals correctly. The data for these plots are artificially generated for illustration purposes.

@@

\subsection{Unsettled practices}
\tocsubsubsectioncomment{Include discussion of active learning/ fast ABFE for virtual screening?}

@@

Best practices for reporting metrics are also evolving. The metric \textit{Best N ligands} has been proposed to help understand if a method is particularly performative within the domain of interest. Christopher Bayly has suggested the use of \textit{Best N ligands} to capture if a method selects the true best molecules within the predicted best molecules, as a metric to reflect how likely you are to synthesise the most potent ligands with a budget allowing for \textit{N}. The metric is a specific case of Boltxmann-Enhanced Discrimination of Receiver Operating Characteristic Curve (BEDROC)\cite{swamidass2009influence}. For purposes of practical application, one likely cares more about your models performance at the higher-activity range of the spectrum vs. across the whole dynamic range; most drug designers would prefer a model that can classify non-/weak-binders and rank binders rather than a method that ranks non-/weak-binders and classifies binders. While the metric usefulness in an applied setting, poorer performance in a lower-ranked compounds could indicate issues in the method and understanding outliers is important for understanding domains-of-applicability or highlighting algorithmic limitations. Additionally, choice of \textit{N} could differ largely for the same target and ligands but different users.
